---
title: "Airplane Crash from 1908 to 2021"
author: "Jaretsy Noriega and Chau Tran"
date: "2023-11-06"
output: [slidy_presentation]
---

```{r setup, include = FALSE }
knitr::opts_chunk$set( echo = FALSE, message = FALSE, warning = FALSE, error = FALSE, quietly = TRUE)
```


```{r , results = 'hide'}
library(readr)
library(tidyverse)
library(stringr)
library(readr)
library(ggplot2)
library(gtable)
library(grid)
library(extrafont)
library(plotly)
library(lubridate)
library(readr)
air <- read_csv("Downloads/Airplane_crashes_dataset_since_1908.csv")

```

## INTRODUCTION 

As you know, in Texas, traffic accidents are something we witness almost every week or every day with our own eyes. However, it is not easy to witness aviation accidents and learn about them. For such a large block of metal to fly into the sky, humans must have put in a lot of effort. Therefore, we are curious what kind of impact and involvement people might have when something that big comes down.

[Link for Airplane Crash and Fatalities dataset](https://www.kaggle.com/datasets/landfallmotto/airplane-crashes-dataset-since-1908)


## DATA DESCRIPTION 

Date: Date of accident, in the format - January 01, 2001

Operator: Airline or operator of the aircraft

AC Type: Aircraft type

Aboard: Total aboard (passengers / crew)

Passengers aboard : Passengers abroad

Crew aboard : Crew abroad

All fatalities : Total fatalities aboard (passengers / crew)

Passenger fatalities: Total Passenger fatalities

Crew fatalities: Total Crew fatalities

Ground: Total killed on the ground

Summary: Brief description of the accident and cause if known

## DATA PREPARATION 



```{r , results = 'hide'}

class(air$date)

```

```{r , results = 'hide'}


air$date <- as.Date(air$date, format = "%B %d, %Y")

air$year <- as.numeric(format(air$date, "%Y"))

air = air %>% mutate( year = air$year)

```


```{r , results = 'hide'}

air$date <- as.Date(air$date, format = "%B %d, %Y")

air$day <- as.numeric(format(air$date, "%d"))

air = air %>% mutate( day = factor(air$day, ordered = T))


```




```{r , results = 'hide'}

air$date <- as.Date(air$date, format = "%B %d, %Y")

air$month <- as.numeric(format(air$date, "%m"))

air = air %>% mutate( month = factor(air$month, ordered = T))


```




```{r , results = 'hide'}


air = subset(air, select = -c(flight_no , route, registration, cn_ln, time))


```



```{r , results = 'hide'}

air = air %>% mutate( country = str_split_fixed(air$location,  ',' , 2)) 

```


```{r , results = 'hide'}


air = air %>% mutate( Country = gsub(".*\\," , "", air$country[, 2])) 

```


When I look at the number of fatalities and the number of passenger aboard, I think that I could obtain
the number of survivals by subtracting the number of fatalities from number of passengers aboard. 
Therefore, I subtract the values in the Fatalities column from the values in the Aboard column and then add
those new values to my data set and name it Survival. 

```{r, echo = TRUE}

air = air %>% mutate(Survival = all_aboard - all_fatalities)

```


```{r , results = 'hide'}

air$crew_aboard = as.numeric(air$crew_aboard)
air$crew_fatalities = as.numeric(air$crew_fatalities)

air = air %>% mutate(`Crew Fatalities Ratio` = crew_fatalities/ crew_aboard)


```


```{r , results = 'hide'}

air$passengers_aboard = as.numeric(air$passengers_aboard)
air$passenger_fatalities = as.numeric(air$passenger_fatalities)

air = air %>% mutate(`Passenger Fatalities Ratio` = passenger_fatalities/ passengers_aboard)


```


```{r , results = 'hide'}

air = air %>% mutate(month = recode(month, '1' = 'January', '2' = 'Feburary', '3' = 'March', '4' = 'April', '5' = 'May', '6' = 'June', '7' = 'July', '8' = 'August', '9' = 'September', '10' = 'October', '11' = 'November', '12' = 'December'))

```



## DATA EXPLORATORY ANALYSIS  

Domain question: 

How has our chance to survive an airplane crash or not get into a crash evolved over 113 years (1908-2021) ? 

Other questions:

What are various factors contributing to airplane crashes over 113 years? 

How large is our chance to survive depending on our roles on the plane ? 

How much damage will be caused when an airplane crash and to what extent ? (number of death )

Is there a safe month or day to fly ? 

What type of airplane is the most dangerous to fly ?

What operator is the most dangerous to fly ?

What are the most dangerous countries to fly ? 


## NUMBER OF PASSENGER ON BOARD AND NUMBER OF FATALITIES BY YEARS 

What is the pattern of number of passenger on board and the number of fatalities over the years ? 

```{r}

air = air[!is.na(air$all_aboard), ]
air = air[!is.na(air$all_fatalities), ]


a = air %>% group_by(year) %>% summarise( Fa = sum(all_fatalities))

  
b = air %>% group_by(year) %>% summarise( A = sum(all_aboard))

c = inner_join(a,b)

#colors <- c("Passengers" = "blue"
#p1 = c %>% ggplot( aes( x= year))
#p1 = p1 + geom_line(aes(y = Fa), color = "red")
#p1 = p1 + geom_line(aes(y = A), color = "blue")
#ggplotly(p1)


colors <- c("Passengers" = "blue", "Fatalities" = "red")
p1 = c %>% ggplot( aes( x= year))  + 
  geom_line(aes(y = Fa, color = "Fatalities")) + 
  geom_line(aes(y = A, color = "Passengers")) + 
  labs( x= "Year", y = "Count", color = "Legend") + 
  scale_color_manual(values = colors) + ggtitle("NUMBER OF PASSENGERS ABOARD AND NUMBER OF FATALITIES BY YEAR")

ggplotly(p1)

 



```



## NUMBER OF SURVIVAL AND NUMBER OF FATALITIES BY YEAR 

How does the chance of survival look like over the years ? 

```{r, echo = FALSE, message = FALSE, warning = FALSE, error = FALSE, }




a = air %>% group_by(year) %>% summarise( Fatalities = sum(all_fatalities))
b = air %>% group_by(year) %>% summarise( Survivors = sum(Survival))
c = inner_join(a,b)

#p1 = c %>% ggplot( aes( x= year))
#p1 =p1 + geom_line(aes(y = Fatalities), color = "red")
#p1 = p1 + geom_line(aes(y = Survivors), color = "green")
#ggplotly(p1)

colors <- c("Survivors" = "green", "Fatalities" = "red")

p1 = c %>% ggplot( aes( x= year))  + 
  geom_line(aes(y = Fatalities, color = "Fatalities")) + 
  geom_line(aes(y = Survivors, color = "Survivors")) + 
  labs( x= "Year", y = "Count", color = "Legend") + 
  scale_color_manual(values = colors) + ggtitle("NUMBER OF SURVIVORS AND NUMBER OF FATALITIES BY YEAR")
p1



```



## NUMBER OF SURVIVAL AND NUMBER OF FATALITIES IN YEAR 1999

Let's zoom in year 1999 to see why the number of survivors was greater than the number of fatalities
```{r}


a = air %>% filter(year == 1999) %>% group_by(date) %>%summarise( Survivors = sum(Survival))
b = air %>% filter(year == 1999) %>% group_by(date) %>%summarise( Fatalities = sum(all_fatalities))


c = inner_join(a,b)

#p1 = c %>% ggplot( aes( x= date))
#p1 = p1 + geom_line(aes(y = Fa), color = "red")
#p1 = p1 + geom_line(aes(y = S), color = "green")


colors <- c("Survivors" = "green", "Fatalities" = "red")

p1 = c %>% ggplot( aes( x= date))  + 
  geom_line(aes(y = Fatalities, color = "Fatalities")) + 
  geom_line(aes(y = Survivors, color = "Survivors")) + 
  labs( x= "Year", y = "Count", color = "Legend") + 
  scale_color_manual(values = colors) + ggtitle("NUMBER OF SURVIVORS AND NUMBER OF FATALITIES IN 1999")

ggplotly(p1)
```

## NUMBER OF PASSENGERS ABOARD AND NUMBER OF FATALITIES 

What pattern of damage does an airplane usually cause ? 

```{r}

b = air %>% ggplot() + aes(x = all_fatalities, y = all_aboard) + geom_point() + xlab("Fatalities") + ylab("Passengers")+
  ggtitle("NUMBER OF PASSENGERS ABOARD AND NUMBER OF FATALITIES ")

ggplotly(b)

```

## NUMBER OF SURVIVORS AND NUMBER OF FATALITIES 

Let's zoom in another perspective to see the pattern of the way an airplane crashes cost lives better! 


```{r}

s = air %>% ggplot() + aes(x = all_fatalities, y = Survival) + geom_point() + xlab("Fatalities") + ylab("Survivors") +
  ggtitle("NUMBER OF SURVIVORS AND NUMBER OF FATALITIES")

ggplotly(s)
```



## NUMBER OF CRASHES BY YEAR 

What time periods have the most crash ? 

```{r}

n = air %>% group_by(year) %>% summarise( Crashes = n()) %>% 
  ggplot() + aes(x = year, y = Crashes)  + geom_line() + xlab("Year") +
  ggtitle("NUMBER OF CRASHES BY YEAR")
ggplotly(n)

```


## NUMBER OF CRASHES BY OPERATORS 

What operators have the most crashes ? 

```{r}

air %>% group_by(operator) %>% summarise( Crashes = n()) %>% arrange(-Crashes) %>% head(10) %>% 
  ggplot() + aes(x = reorder(operator, +Crashes), y = Crashes, fill = operator) + geom_col() + coord_flip() + xlab("Operator") + 
  ggtitle("NUMBER OF CRASHES BY OPERATOR")

```



## NUMBER OF CRASHES BY TYPE 

What type of airplane has the most crashes ? 

```{r}
air %>% group_by(ac_type) %>% summarise( Crashes = n()) %>% arrange(-Crashes) %>% head(10) %>% 
  ggplot(aes(x = reorder(ac_type, +Crashes), y = Crashes, fill = ac_type))   + geom_col() + coord_flip() + ggtitle("NUMBER OF CRASHES BY TYPE") + xlab("Type") 

```


## THE NUMBER OF FATALITIES ON BOARD AND THE NUMBER OF FATALITIES ON GROUND BY YEAR 

How does the number of fatalities on board and the number of fatalities on the ground look like over the years ? 

```{r}
air = air[!is.na(air$all_fatalities), ]
air = air[!is.na(air$ground), ]

a = air %>% group_by(year) %>% summarise( Fatalities = sum(all_fatalities))
air$ground = as.numeric(air$ground)
  
b = air %>% group_by(year) %>% summarise( Ground = sum(ground))

c = inner_join(a,b)



#p1 = c %>% ggplot( aes( x= year))
#p1 = p1 + geom_line(aes(y = Fatalities), color = "red")
#p1 = p1 + geom_line(aes(y = Ground), color = "black")
#ggplotly(p1)

colors <- c("Fatalities on the ground" = "black", "Fatalities aboard" = "red")

p1 = c %>% ggplot( aes( x= year))  + 
  geom_line(aes(y = Fatalities, color = "Fatalities aboard")) + 
  geom_line(aes(y = Ground, color = "Fatalities on the ground")) + 
  labs( x= "Year", y = "Count", color = "Legend") + 
  scale_color_manual(values = colors) + 
  labs(title = "NUMBER OF FATALITIES ABOARD AND \nNUMBER OF FATALITIES ON GROUND BY YEAR")

p1

```



## NUMBER OF CRASHES BY MONTH 

What month has the most crashes ? 

```{r}

air = air[!is.na(air$month), ]
air %>% group_by(month) %>% summarise( Crashes = n()) %>% 
  ggplot() + aes(x = month, y = Crashes, fill = month)  + geom_col() + xlab("Month") + ggtitle("NUMBER OF CRASHES BY MONTH") + 
  theme(axis.text.x = element_text(angle = 60, hjust =1 ))
 
```

## NUMBER OF CRASHES BY DAY 

Does the chance of getting into an airplane crash vary on different days ? 

```{r}

air = air[!is.na(air$day), ]

air %>% group_by(day) %>% summarise( Crashes = n()) %>% 
  ggplot() + aes(x = day, y = Crashes, fill = day)  + geom_col()  + xlab("Day") + ggtitle("NUMBER OF CRASHES BY DAY")  


```

## NUMBER OF CRASHES BY COUNTRY

What countries have the most airplane crashes ? 

```{r}

air = air[-which(air$Country == ""), ]

p1 = air %>% group_by(Country) %>% summarise( Crashes = n()) %>% arrange(-Crashes) %>% head(10) %>% 
  ggplot() + aes( x = reorder(Country, +Crashes), y = Crashes, fill = Country)  + geom_col()  + xlab("Country") +
  ggtitle("NUMBER OF CRASHES BY COUNTRY") 


ggplotly(p1)
```

## RATIO OF FATALITIES OF CREW AND RATIO OF FATALITIES OF PASSENGER BY YEAR 

Does the chance of survival vary on different roles on the airplane ? 

```{r}
air = air[!is.na(air$`Crew Fatalities Ratio`), ]
air = air[!is.na(air$`Passenger Fatalities Ratio`), ]


a = air %>% group_by(year) %>% summarise( Crew = sum(`Crew Fatalities Ratio`))

  
b = air %>% group_by(year) %>% summarise( Passenger = sum(`Passenger Fatalities Ratio`))

c = inner_join(a,b)


#colors = c("Crew" = "orange", "Passenger" = "purple")

#p1 = c %>% ggplot( aes( x= year))
#p1 = p1 + geom_line(aes(y = Crew), color = "Crew")
#p1 = p1 + geom_line(aes(y = Passenger), color = "Passenger")
#p1  + scale_color_manual(values = colors)


#p1 = c %>% ggplot( aes( x= year))
#p1 = p1 + geom_line(aes(y = Crew), color = "orange")
#p1 = p1 + geom_line(aes(y = Passenger), color = "purple")



colors <- c("Passengers" = "purple", "Crews" = "orange")
p1 = c %>% ggplot( aes( x= year))  + 
  geom_line(aes(y = Crew, color = "Crews")) + 
  geom_line(aes(y = Passenger, color = "Passengers")) + 
  labs( x= "Year", y = "Count", color = "Legend") + 
  scale_color_manual(values = colors) + ggtitle("RATIO OF FATALITIES OF CREW VS RATIO OF FATALITIES OF PASSENGERS \nBY YEAR")

p1

```


## SUMMARY

Most of the time, when a plane crashed, passengers on board either mostly died or mostly survived. The survival rate of the passengers on board in 1908-2021 has been relatively low. However, deaths on the ground caused by plane crashes have almost never occurred. But once that happens, the plane crash had the potential to kill thousands of people on the ground like the 9/11 event. We also know that the three most dangerous countries to fly in are Russia, the United States, and Brazil. However, this might be suspected because those countries might have high air traffic so we cannot not conclude yet. In terms of the time of year to fly, the chances of being involved in a plane crash are fairly similar across all months and days of the year. On top of that, regardless of whether a person on a flight is a captain or a passenger, the chances of survival remain the same. The time with the most plane crashes is during the war and the type of plane involved in the most crashes (Douglas DC - 3) is the type used in the war. We have not found any commercial type of airplane that has an outstanding high number of crashes. In short, other than the time of war and the points we have not yet verified, air travel looks quite safe after 1999 to the present.





